An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-18 latex_files/ch4/4_2_ma2.tex
updated most content in chapter 2,3
standarized ma2 example

The second example is the second-order moving average (MA2) which is
used by the ELFI package as a fundamental model for testing all
inference implementations. This example is chosen to confirm that our
implementation of the ROMC approach produces sensible results in a
general model, since the previous ones where artificially created by
us.

\subsubsection*{Problem definition}

The second-order moving average (MA2) is a common model used for
univariate time series analysis. The observation at time $t$ is given by:

@@

\noindent
The random variables $w_{k} \sim \mathcal{N}(0,1), k \in \mathbb{Z}$
represent an independent and identically distributed white noise and
$\theta_1, \theta_2$ the dependence from the previous
observations. The number of consecutive observations $T$ is a
hyper-parameter of the model; in our case we will set
$T=100$. Generating an MA2 time-series is pretty easy and efficient
using a simulator, therefore using a likelihood-free inference method
quite convenient. At the specific example, we use the prior proposed
by \cite{Marin2012} for guaranteeing that the inference problem is
identifiable, i.e.\ loosely speaking the likelihood will have just one
mode. The multivariate prior, which is given in the equation
\eqref{eq:ma2_prior}, follows a triangular shape as plotted in figure
\ref{fig:ma2_1}.

@@

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_prior_samples.png}
    \end{center}
    \caption[MA2 example, prior distribution.]{Prior distribution proposed by \cite{Marin2012}. The
      samples follow a triangular shape.}
  \label{fig:ma2_1}
\end{figure}

\noindent
The vector $\yb_0 = (y_1, \ldots, y_{100})$ used as the observation,
has been generated with $\thetab=(0.6, 0.2)$. The dimensionality of
the output $\yb$ is quite large, therefore we use summary
statistics. Considering that the ouptput vector represents a
time-series signal, we prefer the autocovariance as the summary
statistic; we incorporate the autocovariances with $lag=1$ and
$lag=2$, as shown in equation \eqref{eq:ma2_summary}. Hence, the
distance is defined as the squared euclidean distance between the
summary statistics; this is the same choice as in \cite{Marin2012}.

\begin{gather} \label{eq:ma2_summary}
  s_1(\yb) = \frac{1}{T-1} \sum_{t=2}^T y_ty_{t-1}\\
  s_2(\yb) = \frac{1}{T-2} \sum_{t=3}^T y_ty_{t-2} \\
  s(\yb) = (s_1(\yb), s_2(\yb))\\
  d = ||s(\yb) - s(\yb_0)||_2^2
\end{gather}

@@

As in the previous example, we perform the inference using the two
optimisation alternatives, the gradient-based and the Bayesian
optimisation. In this way, we compare the results obtained in each
step. For comparing the samples drawn with our implementation, we will
use the samples obtained with Rejection ABC. In figure \ref{fig:ma2_2}
we observe that in most cases the optimal distance $d_i^*$ is close to
zero in both optimisation approaches.

In figure \ref{fig:ma2_5}, we have chosen three different
deterministic optimisation cases (i.e. three different seeds) for
illustrating three different cases. In the first case, both
optimisation schemes lead to the creation of almost the same bounding
box. In the second case, the bounding box has a similar shape, though
different size and it is shifted along the $\theta_1$ axis. Finally,
in the third case, the bounding boxes are completely different. We can
thus conclude that although fitting a surrogate model has important
computational advantages, there is no guarantee that it will reproduce
accurately the local region around the optimal point. This
approximation error may lead to the construction of a considerably
different proposal region, which in turn, explains the differences in
the histogram of the marginal distributions presented in figure
\ref{fig:ma2_3} and in the approximate posteriors in figure
\ref{fig:ma2_4}.

In figure \ref{fig:ma2_3}, we demonstrate the histograms of the
marginal posteriors, using the three different methods; (a) Rejection
ABC (first line), (b) ROMC with gradient-based optimisation (second
line) and (c) ROMC with Bayesian optimisation (third
line). Undoubtedly, we observe that there is a significant similarity
between the three approaches. The Rejection ABC inference has been set
to infer 10000 accepted samples, with threshold $\epsilon=0.1$. The
large number of samples and the small distance from the observations
let us treat them as ground-truth information. In the table \ref{tab:ma2}
we present the empirical mean $\mu$ and standard deviation $\sigma$
for each inference approach. We observe that there is a significant
agreement between the approaches, which verifies that the ROMC
implementation provides sensible results. Finally, in figure
\ref{fig:ma2_4} we provide the ROMC approximate posteriors using
gradients and Bayesian optimisation; as confirmed by the statistics in
\ref{tab:ma2}, both posteriors have a single mode, located at the same
point, with a larger variance observed in the Bayesian Optimisation
case. The posteriors mode is quite close to the parameter
configuration that generated the data i.e.\ $\thetab = (0.6, 0.2)$.

\begin{center} \label{tab:ma2}
\begin{tabular}{ c|c|c|c|c }
\hline
& $\mu_{\theta_1}$ & $\sigma_{\theta_1}$ & $\mu_{\theta_2}$ & $\sigma_{\theta_2}$ \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.495 & 0.136 & 0.048 & 0.178 \\
\hline
ROMC (Bayesian optimisation) & 0.510 & 0.156 & 0.108 & 0.183 \\
\hline
\end{tabular}
\end{center}

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist_bo.png}
    \end{center}
    \caption[MA2 example, histogram of distances]{Histogram of distances
      $d_i^*, i \in \{ 1, \ldots, n_1 \}$. The left graph corresponds
      to the gradient-based approach and the right one to the Bayesian
      optimisation approach.}
  \label{fig:ma2_2}
\end{figure}

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_1.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_1_bo.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_2.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_2_bo.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[MA2 example, the acceptance region in three distinct optimisation problems.]{Visualisation of the acceptance region in 3 different optimisation problems. Each row illustrates a different optimisation problem, the left column corresponds to the gradient-based approach and the right column to the Bayesian optimisation approach. The examples have been chosen to illustrate three different cases; in the first case, both optimisation schemes lead to similar optimal point and bounding box, in the second case the bounding box is similar in shape but a little bit shifted to the right relatively to the gradient-based approach and in the third case, both the optimal point and the bounding box is completely different.}
  \label{fig:ma2_5}
\end{figure}

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_rejection.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_rejection.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.png}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal distributions.]{Histogram of the marginal distributions using
      different inference approaches; (a) in the first row, the
      approximate posterior samples are obtained using Rejection ABC
      sampling (b) in the second row, using ROMC sampling with
      gradient-based approach and (c) in the third row, using ROMC
      sampling with Bayesian optimisation approach. The vertical (red)
      line represents the samples mean $\mu$ and the horizontal
      (black) the standard deviation $\sigma$.}
  \label{fig:ma2_3}
\end{figure}

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_romc_posterior.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_romc_posterior_bo.png}
    \end{center}
    \caption[MA2 example, approximate posterior.]{ROMC approximate posteriors using gradient-based approach
      (left) and Bayesian optimisation approach (right).}
  \label{fig:ma2_4}
\end{figure}

>>>>

Second-order moving average (MA2) is a probabilistic model commonly
used for univariate time series analysis. The observation at time $t$
is given by:

@@

\noindent
The random variable $w_{k} \sim \mathcal{N}(0,1), k \in \mathbb{Z}$
represent white noise and $\theta_1, \theta_2$, the two parameters of
interest, the dependence from the previous observations. The number of
consecutive observations $T$ is a hyper-parameter of the model that in
our case is set at $T=100$. For guaranteeing that the inference
problem is identifiable, i.e.\ the likelihood has only one mode, we
use the prior proposed by \cite{Marin2012} given in the equation
\eqref{eq:ma2_prior}:

@@

% \begin{figure}[ht]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_prior_samples.png}
%     \end{center}
%     \caption[MA2 example, prior distribution.]{Prior distribution proposed by \cite{Marin2012}. The
%       samples follow a triangular shape.}
%   \label{fig:ma2_1}
% \end{figure}

\noindent
The observation vector $\yb_0 = (y_1, \ldots, y_{100})$ is generated
with $\thetab^*=(0.6, 0.2)$. The dimensionality of the output $\yb$ is
high, therefore we use summary statistics. Considering that the
ouptput vector represents a time-series signal, we choose as summary
statistics the autocovariances with $lag=1$ and $lag=2$, as shown in
equations \eqref{eq:ma2_summary_1} and \eqref{eq:ma2_summary_2}. The
final distance node is the squared euclidean
distance \eqref{eq:ma2_summary_4}.

\begin{gather} \label{eq:ma2_summary_1}
  s_1(\yb) = \frac{1}{T-1} \sum_{t=2}^T y_ty_{t-1}\\ \label{eq:ma2_summary_2}
  s_2(\yb) = \frac{1}{T-2} \sum_{t=3}^T y_ty_{t-2} \\
  s(\yb) = (s_1(\yb), s_2(\yb))\\ \label{eq:ma2_summary_4}
  d = ||s(\yb) - s(\yb_0)||_2^2 
\end{gather}

@@

As in the previous example, we perform the inference using both
optimisation alternatives (gradient-based and Bayesian optimisation)
and we compare the results at each step. In the end, we use the
Rejection ABC algorithm for evaluating both approximations. Rejection
ABC is a robust method in terms of accuracy and can be used for
baseline comparison.

In figure \ref{fig:ma2_5}, we illustrate the acceptance region of the
same deterministic simulator, i.e. with \code{seed=799981517}, in the
gradient-based and the Bayesian optimisation case. The acceptance
regions are similar in both cases. The optimal points are close and
the search directions are similar. This indicates that both
optimization scemes work robustly. Therefore, we observe that as we
move away from the optimal pint, the surrogate model fitted by the
Bayesian Optimisation tends to deviate from the real distance. This
divergence affects the bounding box approximation and explains some
small differences between the posterior approximations, as shown in
figure \ref{fig:ma2_3}. In figure \ref{fig:ma2_3}, we demonstrate the
histograms of the marginal posteriors, using the three different
methods; (a) Rejection ABC (first line), (b) ROMC with gradient-based
optimisation (second line) and (c) ROMC with Bayesian optimisation
(third line). Undoubtedly, we observe that there is a significant
similarity between the three approaches. The Rejection ABC inference
has been set to infer 10000 accepted samples, with threshold
$\epsilon=0.1$. At table \ref{tab:ma2} we present the empirical mean
$\mu$ and standard deviation $\sigma$ for each inference approach, for
verifying their agreement.

\begin{center} \label{tab:ma2}
\begin{tabular}{ c|c|c|c|c }
\hline
& $\mu_{\theta_1}$ & $\sigma_{\theta_1}$ & $\mu_{\theta_2}$ & $\sigma_{\theta_2}$ \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.504 & 0.141 & 0.031 & 0.172 \\
\hline
ROMC (Bayesian optimisation) & 0.5 & 0.165 & 0.07 & 0.167 \\
\hline
\end{tabular}
\end{center}

% \begin{figure}[H]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist.png}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist_bo.png} \\
%     \end{center}
%     \caption[MA2 example, histogram of distances]{Histogram of distances
%       $d_i^*, i \in \{ 1, \ldots, n_1 \}$. The left graph corresponds
%       to the gradient-based approach and the right one to the Bayesian
%       optimisation approach.}
%   \label{fig:ma2_2}
% \end{figure}

\begin{figure}[H]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesia Optimisation.}
  \label{fig:ma2_5}
\end{figure}

\begin{figure}[H]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_rejection.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_rejection.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc.png}\\
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t1_romc_bo.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_hist_t2_romc_bo.png}\\
    \end{center}
    \caption[MA2 example, evaluation of the marginal
    distributions.]{Histogram of the marginal posterior distributions
      using three different inference approaches; (a) in the first
      row, the samples are obtained using Rejection ABC sampling (b)
      in the second row, using ROMC with a gradient-based optimiser
      and (c) in the third row, using ROMC with Bayesian optimization
      approach. The vertical (red) line represents the samples mean
      $\mu$ and the horizontal (black) the standard deviation
      $\sigma$.}
  \label{fig:ma2_3}
\end{figure}

% \begin{figure}[H]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_romc_posterior.png}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_romc_posterior_bo.png}
%     \end{center}
%     \caption[MA2 example, approximate posterior.]{ROMC approximate posteriors using gradient-based approach
%       (left) and Bayesian optimisation approach (right).}
%   \label{fig:ma2_4}
% \end{figure}
